Make manuscript figures

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
add trait value, qq-plot, manhattan plot figures

Load libraries
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(POUMM)
library(ape)
library(cowplot)
library(treeio)

@@

gv_name <- "Pathogen part of trait"
gh_name <- "Host part of trait"
e_name <- "Non-pathogen part of trait"
gvhat_name <- "Estimated pathogen part of trait"
ehat_name <- "Estimated non-pathogen part of trait"
z_minus_zbar_name <- "Scaled trait value"
```

Load data
```{r}
accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
tree <- treeio::full_join(tree, traits, by = "label")
POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv"))
```

@@

@@

Manhattan plot of GWAS results
```{r}

Q-Q plot of GWAS results
```{r}

>>>>

Load libraries
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(POUMM)
library(ape)
library(cowplot)
library(treeio)
library(ggExtra)

@@

gv_name <- "Pathogen part of trait"
gh_name <- "Host part of trait"
e_name <- "Non-pathogen part of trait"
gvhat_name <- "Estimated pathogen part of trait"
ehat_name <- "Estimated non-pathogen part of trait"
z_minus_zbar_name <- "Scaled trait value"
standard_gwas_name <- "GWAS with standard trait value"
corrected_gwas_name <- "GWAS with estimated non-pathogen part of trait"
genome_wide_significance <- 5E-8
```

Load data
```{r}
accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
tree <- treeio::full_join(tree, traits, by = "label")
POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"))
gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
```

@@

Estimated trait values
TODO: think about some way of depicting phylogenetic relationships here too?
```{r}
p <- ggplot(data = traits, aes(x = trait, y = h.MWA)) +
      geom_point() +
      shared_theme +
      labs(x = "Calculated spVL trait", y = ehat_name)
p2 <- ggExtra::ggMarginal(p, type="histogram")
ggsave(plot = p2, filename = paste(FIGDIR, "trait_values.png", sep = "/"),
      width = linewidth * 2/3, height = height * 2/3, units = "cm")
```

@@

Q-Q plot of GWAS results
```{r}
qq_data_standard <- data.frame(
        observed_log10p = -log10(sort(gwas_results$P_standard, decreasing = F)),
        expected_log10p = -log10(ppoints(nrow(gwas_results))),
        gwas_type = standard_gwas_name)
qq_data_phylo <- data.frame(
        observed_log10p = -log10(sort(gwas_results$P_corrected, decreasing = F)),
        expected_log10p = -log10(ppoints(nrow(gwas_results))),
        gwas_type = corrected_gwas_name)
qq_results <- rbind(qq_data_standard, qq_data_phylo)

p <- ggplot(data = qq_results, aes(x = expected_log10p, y = observed_log10p)) +
        geom_point() +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                    color = RColorBrewer::brewer.pal(n = 3, name = diverging_palette)[[1]]) +
        facet_wrap(. ~ gwas_type) +
        labs(x = expression("expected log"[10]*"(p)"), y = expression("observed log"[10]*"(p)")) +
        shared_theme
ggsave(plot = p, filename = paste(FIGDIR, "qq_plots.png", sep = "/"),
       width = linewidth, height = height * 2/3, units = "cm")
```

Manhattan plot of GWAS results
```{r}
# Add genome position to data
chr_lengths <- gwas_results %>%
        group_by(CHROM) %>%
        summarise(chr_len = max(POS))
chr_starts <- chr_lengths %>%
        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
        select(-chr_len)
gwas_results_annotated <- left_join(gwas_results, chr_starts, by = "CHROM")
gwas_results_annotated <- gwas_results_annotated %>%
        arrange(CHROM, POS) %>%
        mutate(BP_position = CHR_start + POS)

# This is a super hacky way to get labels in between x ticks
# Hadley Wickam disapproves :/
max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 22, "BP_position"])
x_axis_ticks <- chr_starts
x_axis_ticks <- rbind(
        x_axis_ticks,
        data.frame(
                CHROM = c(23, 24),
                CHR_start = c(max_BP_position, max_BP_position + 1000)))
x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
x_axis_ticks[x_axis_ticks$CHROM == 22, "midpoint"] <-
        x_axis_ticks[x_axis_ticks$CHROM == 22, "CHR_start"] +
                (max(gwas_results_annotated[gwas_results_annotated$CHROM == 22, "BP_position"]) -
                        x_axis_ticks[x_axis_ticks$CHROM == 22, "CHR_start"]) / 2
x_axis_ticks_tmp <- tidyr::pivot_longer(
        data = x_axis_ticks,
        cols = c("midpoint", "CHR_start"),
        names_to = "marker",
        values_to = "tick_position") %>%
        mutate(CHROM = as.character(as.integer(CHROM)))
x_axis_ticks_tmp[seq(from = 2, to = 46, by = 2), "CHROM"] <- ""
x_axis_ticks_tmp[x_axis_ticks_tmp$CHROM == "23", "CHROM"] <- ""
x_axis_ticks_tmp[x_axis_ticks_tmp$CHROM == "24", "CHROM"] <- ""

# Format data for plotting
results_long <- tidyr::pivot_longer(
        data = gwas_results_annotated,
        cols = c("P_standard", "P_corrected"),
        names_to = "gwas_type",
        names_prefix = "P_",
        values_to = "P")

# Fix facet labels for plotting
results_long$gwas_type <- factor(
        results_long$gwas_type,
        levels = c("standard", "corrected"),
        labels = c(standard_gwas_name, corrected_gwas_name))

# Make comparative GWAS overview plot
overview_comparison <- ggplot(
        data = results_long,
        aes(x = BP_position, y = -log10(P))) +
        geom_point(aes(color = CHROM %% 2 == 0)) +
        scale_color_manual(values = c("grey", "black")) +
        scale_x_continuous(
                breaks = x_axis_ticks$tick_position,
                labels = x_axis_ticks$CHROM,
                limits = c(0, NA),
                expand = c(0, 0)) +
        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
        labs(x = "Chromosome", y = expression('-Log'[10]*'(p)')) +
        geom_hline(
                yintercept = -log10(genome_wide_significance),
                linetype = "dotted") +
        theme_bw() +
        theme(
                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
                panel.grid.major.x = element_blank(),
                axis.ticks.x = element_line(color = rep(c(NA, "black"), 23)),
                panel.grid.minor.x = element_blank(),
                legend.position = "none",
                plot.margin = unit(c(5.5, 5.5, 25, 5.5), "pt"),
                axis.text.x = element_text(size = 6)) +
        facet_grid(. ~ gwas_type)

# Make plot showing p-value changes in the CCR5 and MHC regions
regional_results<- gwas_results_annotated %>% filter(
        (CHROM == 3 & POS >= 45.5E6 & POS <= 47E6) | (CHROM == 6 & POS >= 29.5E6 & POS <= 33.5E6))

# Fix facet labels for plotting
regional_results$CHROM <- factor(
        regional_results$CHROM,
        levels = c("3", "6"),
        labels = c("italic(CCR5)*' region'", "'MHC region'"))

p_vals_regional <- ggplot(
        data = regional_results,
        aes(x = POS/1E6, y = -log10(P_corrected))) +
        geom_point(aes(color = -neg_log10_P_standard_minus_corrected)) +
        scale_color_gradient2(
                name = expression(Delta*'[-Log'[10]*'(p)]'),
                low = "#B2182B",
                mid = "#FFFFBF",
                high = "#2166AC",
                midpoint = 0) +
        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
        labs(
                x = "Base position (Mb)",
                y = expression('-Log'[10]*'(p)')) +
        geom_hline(
                yintercept = -log10(genome_wide_significance),
                linetype = "dotted") +
        theme_bw() +
        theme(
                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
                panel.grid.major.x = element_blank(),
                axis.ticks.x = element_line(color = rep(c(NA, "black"), 23)),
                panel.grid.minor.x = element_blank(),
                plot.margin = unit(c(25, 5.5, 5.5, 5.5), "pt")) +
        facet_grid(. ~ CHROM, labeller = label_parsed, scales = "free")

# Assemble the plots as subplots
png(
        filename = paste(FIGDIR, "gwas_results.png", sep = "/"),
        width = linewidth, height = height, unit = "cm", res = 300)
gridExtra::grid.arrange(
        overview_comparison,
        p_vals_regional,
        nrow = 2, heights = c(1.5, 2))
dev.off()
```

Debugging
```{r}
# Are trait values very different from what was inferred last time?
old_traits <- read.delim(file = "/Users/nadeaus/Polybox/200207_phenotypes.txt", sep = "", header = T)
trait_comparison <- merge(
        x = old_traits, y = traits,
        by.x = "IID", by.y = "id",
        all = T)

ggplot(data = trait_comparison, aes(x = total_spvl, y = trait)) + geom_point()  # calculated spVL indistinguishable
ggplot(data = trait_comparison, aes(x = corrected_spvl, y = h.MWA)) + geom_point()  # corrected spVL very similar, well correlated

# What about GWAS phenotypes file?
gwas_phenotypes <- read.delim(file = "output/gwas_phenotypes.txt", sep = "", header = T)
gwas_pheno_comparison <- merge(
        x = old_traits, y = gwas_phenotypes,
        by = "IID",
        all = T)
ggplot(data = gwas_pheno_comparison, aes(x = corrected_spvl, y = h.MWA)) + geom_point()  # also well correlated but previous GWAS had more samples?
```
